|

### 2022-11-09 ms/ms.Rmd
<edit ms>

As noted already, this prevalence was remarked on almost 50 years earlier, the result of large-scale collections acquisitions that major museums had engaged in since the late 1800s (e.g., Květinová 2011:65-66).  But Chimú pottery continued to flow into established museum collections, and to be sought as new museums were established (see Mowat 1988 for an example of donation and accessioning practices).  Collection-building accelerated as new museums were founded after World War II  and continued with limited constraints until after 1970, when the UNESCO Convention regarding cultural property established guidelines that gradually altered legal standards and ethical practices surrounding collecting (Gerstenblith 2013).  Although some museums continued to send curators to Peru to purchase from antiquities dealers, a cursory review of museum records reveals that many new accessions came as gifts from individual donors, who provided little or no provenance information.  For example, the Metropolitan Museum of Art received a feline bottle (64.228.17) from Nathan Cummings in 1964, who had purchased the vessel a decade earlier from a collector in Buenos Aires.   The piece was one of more than 1000 artifacts that Cummings is credited with donating to the museum.  The Cleveland Museum of Art, founded in 1913, also built its collections through donations, including a blackware jar (1959.333) given by William Ellery Greene in 1959.   It was the only Andean artifact given to the museum by the donor.

Building on preliminary work by Pedro Rojas Ponce and Dorothy Menzel, Sheele and Patterson (1966) compared illustrations from the published literature with vessels from Peru’s Museo Nacional de Antropología y Arqueología and Harvard University’s Peabody Museum of Archaeology and Ethnology.  They created a preliminary seriation of Chimú ceramics from the end of the Middle Horizon to the early Colonial Period, distinguishing seven different phases based on the identification of different vessel shapes, decorative elements, and firing characteristics.  Although the seriation was developed to build a more fine-grained relative chronology, the small sample sizes of some phases presented interpretive challenges.  For example, the fourth phase _Lambayeque_, was based on seven vessels from a single grave in the Lambayeque Valley, located approximately 175 km to the north of Chan Chan, and probably reflects regional ceramic diversity more than a distinct period of Chimú production in the Moche Valley.  Other phases were based on small sample sizes and vessels with limited provenance information.  One important distinction in the Scheele and Patterson seriation was between Chimú ceramics of the late LIP (Chimu Phase T-1) and those from the time of the Inca occupation (Chimu Phase T-2).  The former was defined based on the prevalence of stirrup-spout vessels and the _less carefully made_ production of mold-produced vessels that seemed to represent a more restricted set of shapes than the preceding phase (Trujillo Phase T-2).  The Chimú-Inca phase maintained elements of existing ceramic production, adding features from Inca vessel shapes.  Scheele and Patterson note (1966:24) that this phase seems to be distributed beyond earlier Chimu political boundaries and economic networks—it has been identified on the south and central coast and in the Cuzco region.  Overall, Scheele and Patterson’s work clarifies how some vessel shapes developed over time, making it possible to assign provenienced gravelots to more specific late prehispanic periods.  It also identified some important aspects of production practices and hybridity that have been important for subsequent archaeological fieldwork.

@@

Although the first attempts at defining Chimú ceramics within a broader north coast sequence utilized museum collections, archaeological excavations have increasingly contributed to the reconstruction of ceramic production practices.  In the mid-20th century, excavators turned to cemeteries as a source of well-provenienced vessels that could be used to elaborate the regional sequence (e.g., Collier 1955; Willey 1947).  Such work represented only a miniscule proportion of Chimú ceramics that were being dug up at that time, most of which were obtained by huaqueros whose illicit work supplied private collectors and museums around the world.  Studies of whole vessels concluded that the use of molds increased in late prehispanic periods on the north coast (Collier 1955), and by the 1960s, some researchers had identified molds and matrices in the pottery collections acquired by some museums (e.g., Grossman 1969-1970; Thompson 1963).  Such work constituted an important step toward moving from straightforward aesthetic considerations to questions about the production and consumption of Chimú ceramics, but it remained rooted in mortuary collections that did not focus on evidence of production practices.

At the end of the 1960s, Moseley and Mackey’s Chan Chan-Moche Valley Project (1969-1975) began to increase the intensity of professional archaeological research, with regional survey work in the lower Moche Valley and mapping and excavations at Chan Chan and other nearby sites.  While not focused on cemetery sites, this project and related studies encountered additional mortuary contexts with well-preserved ceramics (see Donnan and Mackey 1978).  It was not until the 1990s, however, that archaeologists began to identify loci of ceramic production based on the presence of overfired sherds, molds and matrices, stamps, tools, kilns, and other evidence (Donnan 1997; Hayashida 1999; Mackey 2003; Mackey and Sapp 2021; Tschauner 2006, 2009).  Some of the workshops reflect ceramic production practices during the preceding era of Chimú imperial expansion, whereas others were in use during the Inca occupation of the north coast.  Surface collections and excavations in production areas generated multiple lines of material evidence, including large numbers of potsherds that could not easily be assigned to the diverse array of shapes identified in collection-based classifications.

The study of Chimú and Chimú-Inca pottery workshops has focused on two central themes:  specialization and cultural (dis)continuity under foreign rule.  These issues were nascent in many of the earlier collection-based studies, but the careful excavation and analysis of production contexts added invaluable evidence to the discussion.  Molds and matrices encountered at workshops indicate that multiple vessel parts were mass-produced and had to be assembled following a specific sequence, which Tschauner (2006:179) interprets as evidence of specialized production during Chimú times.  Tschauner’s excavations at the Pampa de Burros workshop in the Lambayeque Valley identified heavy use of vertical half-molds to produce jars, canteens or flasks, bottles, and ollas—virtually all of the shapes that were used by Chimú-era settlements in the surrounding valley (2006:182).  He sees the local potters as relatively unskilled producers who relied on more adept makers of ceramic molds, and whose _market-oriented_ work could flexibly adapt to consumer preferences (2006:183-185).  Levine (2011) builds on this idea, noting that some of the variability seen in the Chimú assemblage could come from ceramic producers _mixing and matching_ molds for different vessel parts and adornments to produce new combinations.

For the Inca period, scholars acknowledge a degree of continuity in local potting practices, but also some significant changes.  At Tambo Real and La Viña in the Leche Valley, Hayashida (1999) notes the continuity of local practices for shaping vessels (press molds, paddle and anvil) and firing.  While local jar shapes persisted in Inca times, she observes the presence of a mold for an Inca flared jar neck (1999:345), which could be added to jars that were shaped using local press molds.  Overall, Hayashida sees little evidence of Inca _retraining_ of Chimú potters.

Other Inca-Chimú workshops in the Jequetepeque Valley provide evidence that practices varied at different production locations.  Donnan’s (1997:32-35) surface collections at Cañoncillo encountered dozens of mold fragments, most of which were for shaping the bottom and top hemispheres of ollas.  Donnan identified molds for shaping jars along a vertical axis, and for producing bottles, including the distinctive Inca “aryballoid” shape.  Although he reports that no molds appeared for shaping handles or rims, there were molds for producing stirrup spouts, as well as for humans, birds, and other animals.  The production evidence from this site suggests that while some vessels were largely shaped using molds, some elements were hand-formed and showed variations suggesting the work of multiple potters.  At the Inca administrative center of Farfán, Mackey (2004:336-338) excavated a small workshop that produced large jars (tinajas) for brewing or storage, using the paddle-and-anvil technique.  Finally, the excavation of three workshop patios at El Algarrobal de Moro, a lower-order Chimú administrative center, encountered 25 mold fragments, including molds for a stirrup spout vessel and a miniature jar (Mackey and Sapp 2021:21-22).

@@

Given the acknowledged challenges in classifying Chimú vessels, it is not surprising that recent ceramic studies have embraced technological analyses, including petrography (Krzanowski and Pawilowski 1980), neutron activation, X-ray diffraction, X-ray fluorescence, particle-induced X-ray emission (Cunha Lima 2010), and Mössbauer spectroscopy (Tschauner and Wagner 2003).  Despite such work, Shimada and Wagner (2019) argue that the reconstruction of the chaîne opératoire for north coast blackwares remains incomplete.  It should be noted that many of the techniques used to reconstruct production practices rely on samples of potsherds that can be removed to the laboratory, and potentially altered or destroyed in the process of analysis.  More recently, researchers have brought non-destructive techniques such as CT scans (Wauters 2016) and linear morphometric analysis (Květinová 2011) to the study of museum collections.  This work emphasizes the variability found in different vessel categories, as well as the reconstruction of productive processes, including the use of molds to manufacture vessel bodies and other elements.

@@

These two questions can be addressed using three-dimensional data from well-preserved vessels, an approach that presents logistical challenges.  To date, Wauters (2016) has used CT scans and X-ray radiography to study 16 stirrup-spout bottles from the Royal Museums of Art and History of Brussels and the Ethnography Museum of Geneva.  To produce the scans, the vessels had to be removed from the museums and transferred to hospitals where the necessary equipment could be accessed.  While generating high-resolution data, funding constraints and curatorial policies would make it difficult to conduct such research with large sample sizes.  As an alternative, portable high-resolution 3D scanning equipment can be brought into museums to scan collections in situ, using the resulting data for geometric morphometric analysis.  Between 2018 and 2020, we conducted exploratory research on Chimú ceramics in two collections using this approach.

@@

For our broader study, we scanned #### prehispanic ceramic vessels from two collections that reflect the broader history of collecting and curating Chimú ceramics.  Our first collection was the Art and Art History Collection (AAHC), a modest sized collection that accumulated haphazardly at the University of Texas at Austin as numerous individuals donors gave the university artifacts that had little or no provenance information.  The ancient pottery now in the AAHC was previously curated by the Texas Memorial Museum, which transferred them to the department of Art and Art History in 2004 as the museum implemented new curatorial priorities. In 2018, we scanned 112 Andean vessels from the AAHC collection. The data presented here come from a sample of Chimú bottles (n=23) and jars (n=16) from the UT collection.  _**[Astrid:  How many Andean vessels overall in the collection, and how much Chimú?  Should we include a table of the samples here, with as much provenance info as we can find?]**_

As a second collection, we also worked with a large museum sample that was acquired in Peru by a professional archaeologist the early days of museum collecting.  The Bandelier Collection at the American Museum of Natural History consists of nearly 8,000 Andean artifacts from Peru and Bolivia.  The Swiss-American archaeologist Adolph Bandelier acquired these pieces as part of his 1890s Andean expedition, which was initially funded by Henry Villard.  Bandelier visited the Trujillo area in 1893, where he sketched the ruins of Chan Chan and acquired more than 400 Chimú vessels and other artifacts that were accessioned by the museum.  Although the records of Bandelier’s expedition remain unpublished, the collection offers a large sample size of vessels that were collected at the same time, during a period of archaeological collecting when dealers were less attuned to the targeted collecting that Kroeber observed a generation later. In 2020, we scanned ### Chimú, Inca, and Chimú-Inca ceramic vessels at the AMNH.

The exploratory nature of the work also raised ethical considerations regarding work with museum collections.  Archaeologists have intensified their work with museum collections in recent years, and the consideration of whether to conduct such research are as important as the logistical questions of how to accomplish the work.  The ethical principles of the Society for American Archaeology highlight the values of stewardship and public responsibility—it is important for archaeologists to care for the material record, and also to share information about it with the public—identifying “public education and outreach” as significant.  From this perspective, the non-destructive and museum-based work of 3D scanning presented little risk to the artifacts being analyzed, and our research was designed to be available to the public.  When working with the AAHC vessels, we developed an educational component, including a small class of UT undergraduates in the lab-based collection of linear measurements of the vessels being scanned.  That supervised research provided a diverse group of undergraduates with artifact analysis, while also teaching them about the loss of heritage and scientific knowledge that accompanies the collection of antiquities.

While the careful study and public reporting of museum collections is consistent with professional ethics in Americanist archaeology, the SAA also identifies a potential ethical concern in doing so:  the commercialization of the archaeological record.  The SAA clearly states that “Archaeologists should therefore carefully weigh the benefits to scholarship of a project against the costs of potentially enhancing the commercial value of archaeological objects. Whenever possible they should discourage, and should themselves avoid, activities that enhance the commercial value of archaeological objects, especially objects that are not curated in public institutions, or readily available for scientific study, public interpretation, and display.”   As described above, Chimú blackware pottery is an artifact class that was heavily collected in the past, and it is openly sold today by galleries and internet vendors.  As Atwood (2004:22-23) notes, huaqueros, dealers, and collectors are no longer interested in trafficking Andean ceramics, especially mass-produced wares like those of the Chimú Empire.  Our work with publicly-held collections poses little risk of increasing the present commercial value of Chimú ceramics, while its public discussion of the losses that accompany undocumented excavations and illicit collection constitutes a benefit that outweighs any associated risks.

@@

Linear shape variables... (Figure \ref{fig:linear}).

@@

Following transfer to Dx, each mesh will be subjected to an additional quality check to eliminate non-manifold poly-vertices, folded poly-faces, dangling poly-faces, small clusters, small poly-faces, non-manifold poly-faces, crossing poly-faces, and small tunnels. Due to the paucity of homologous landmarks on cultural artefacts \citep{RN8521}, reference geometry will be constructed around each vessel in a manner that yields a replicable configuration of nine landmarks, and 46 equidistant semilandmarks along the widest vessel profile, with notable similarities to previous landmark configurations \citep{RN7039,RN7925,RN8071,RN8361,RN8967,RN8375}, which largely follow \citet{RN5700}.

@@

Landmark data were aligned to a global coordinate system \citep{RN8477,RN7502,RN11622,RN11623,RN11563}, achieved through generalised Procrustes superimposition \citep{RN11138,RN478,RN1646} in R using the `geomorph` and `RRPP` packages \citep{RN1655,RN11775,RN11530,RN1774,RN9565}. Procrustes superimposition translates and rotates the coordinate data to allow for comparisons among objects, while also scaling each biface using unit-centroid size---the square root of the sum of squared distances from each landmark to the specimen’s centroid \citep{RN11139,RN11140,RN11564,RN478}. The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space (Figure \ref{fig:gpa}) \citep{RN11141,RN11142,RN1646,RN11563}.

@@

Linear measurements from the sample of jars and bottles included vessel height, rim thickness, and the diameter of each vessel at its orifice, neck, belly, and base.  Of these metrics, vessel height, base diameter, and neck diameter did not differ significantly between the two samples.  Overall, the jars and bottles in our pilot sample overlap in some key dimensions, with the bottles showing greater overall size variations, which would be not be consistent with production using a single set of molds.  The other three sets of linear measurements showed significant differences between the jars and bottles in our sample, in terms of orifice diameter, belly diameter, and rim thickness.  Given the different production scenarios emerging from archaeological field analyses—a combination of hand-shaping and use of whole vessel or elemental molds, the significant differences in orifice and belly diameters might indicate that the kinds of bottles and jars in our sample were produced as distinct shapes, rather than by mixing and matching different molds to cater to changing consumer tastes.  The differences in rim thickness suggest that distinct production practices were used when producing bottles and jars. [_**Zac:  Should we discuss the analysis of the figures on the vessels, and then point out that it’s inconclusive, or just leave that out?**_]

>>>>

As noted already, this prevalence was remarked on almost 50 years earlier, the result of large-scale collections acquisitions that major museums had engaged in since the late 1800s \citep[65-66]{RN11159}.  But Chimú pottery continued to flow into established museum collections, and to be sought as new museums were established (see Mowat 1988 for an example of donation and accessioning practices).  Collection-building accelerated as new museums were founded after World War II  and continued with limited constraints until after 1970, when the UNESCO Convention regarding cultural property established guidelines that gradually altered legal standards and ethical practices surrounding collecting \citep{RN11160}.  Although some museums continued to send curators to Peru to purchase from antiquities dealers, a cursory review of museum records reveals that many new accessions came as gifts from individual donors, who provided little or no provenance information.  For example, the Metropolitan Museum of Art received a feline bottle (64.228.17) from Nathan Cummings in 1964, who had purchased the vessel a decade earlier from a collector in Buenos Aires.   The piece was one of more than 1000 artifacts that Cummings is credited with donating to the museum.  The Cleveland Museum of Art, founded in 1913, also built its collections through donations, including a blackware jar (1959.333) given by William Ellery Greene in 1959.   It was the only Andean artifact given to the museum by the donor.

Building on preliminary work by Pedro Rojas Ponce and Dorothy Menzel, \citet{RN11158} compared illustrations from the published literature with vessels from Peru’s Museo Nacional de Antropología y Arqueología and Harvard University’s Peabody Museum of Archaeology and Ethnology.  They created a preliminary seriation of Chimú ceramics from the end of the Middle Horizon to the early Colonial Period, distinguishing seven different phases based on the identification of different vessel shapes, decorative elements, and firing characteristics.  Although the seriation was developed to build a more fine-grained relative chronology, the small sample sizes of some phases presented interpretive challenges.  For example, the fourth phase _Lambayeque_, was based on seven vessels from a single grave in the Lambayeque Valley, located approximately 175 km to the north of Chan Chan, and probably reflects regional ceramic diversity more than a distinct period of Chimú production in the Moche Valley.  Other phases were based on small sample sizes and vessels with limited provenance information.  One important distinction in the Scheele and Patterson seriation was between Chimú ceramics of the late LIP (Chimu Phase T-1) and those from the time of the Inca occupation (Chimu Phase T-2).  The former was defined based on the prevalence of stirrup-spout vessels and the _less carefully made_ production of mold-produced vessels that seemed to represent a more restricted set of shapes than the preceding phase (Trujillo Phase T-2).  The Chimú-Inca phase maintained elements of existing ceramic production, adding features from Inca vessel shapes. \citet[24]{RN11158} note that this phase seems to be distributed beyond earlier Chimu political boundaries and economic networks---it has been identified on the south and central coast and in the Cuzco region.  Overall, Scheele and Patterson’s work clarifies how some vessel shapes developed over time, making it possible to assign provenienced gravelots to more specific late prehispanic periods.  It also identified some important aspects of production practices and hybridity that have been important for subsequent archaeological fieldwork.

@@

Although the first attempts at defining Chimú ceramics within a broader north coast sequence utilized museum collections, archaeological excavations have increasingly contributed to the reconstruction of ceramic production practices.  In the mid-20th century, excavators turned to cemeteries as a source of well-provenienced vessels that could be used to elaborate the regional sequence \citep{RN8727,RN11161}.  Such work represented only a miniscule proportion of Chimú ceramics that were being dug up at that time, most of which were obtained by huaqueros whose illicit work supplied private collectors and museums around the world.  Studies of whole vessels concluded that the use of molds increased in late prehispanic periods on the north coast \citep{RN8727}, and by the 1960s, some researchers had identified molds and matrices in the pottery collections acquired by some museums \citep{RN11162,RN11163}.  Such work constituted an important step toward moving from straightforward aesthetic considerations to questions about the production and consumption of Chimú ceramics, but it remained rooted in mortuary collections that did not focus on evidence of production practices.

At the end of the 1960s, Moseley and Mackey’s Chan Chan-Moche Valley Project (1969-1975) began to increase the intensity of professional archaeological research, with regional survey work in the lower Moche Valley and mapping and excavations at Chan Chan and other nearby sites.  While not focused on cemetery sites, this project and related studies encountered additional mortuary contexts with well-preserved ceramics \citep{RN11165}.  It was not until the 1990s, however, that archaeologists began to identify loci of ceramic production based on the presence of overfired sherds, molds and matrices, stamps, tools, kilns, and other evidence \citep{RN11164,RN8875,RN11166,RN11167,RN11168,RN11169}.  Some of the workshops reflect ceramic production practices during the preceding era of Chimú imperial expansion, whereas others were in use during the Inca occupation of the north coast.  Surface collections and excavations in production areas generated multiple lines of material evidence, including large numbers of potsherds that could not easily be assigned to the diverse array of shapes identified in collection-based classifications.

The study of Chimú and Chimú-Inca pottery workshops has focused on two central themes:  specialization and cultural (dis)continuity under foreign rule.  These issues were nascent in many of the earlier collection-based studies, but the careful excavation and analysis of production contexts added invaluable evidence to the discussion.  Molds and matrices encountered at workshops indicate that multiple vessel parts were mass-produced and had to be assembled following a specific sequence, which \citet[179]{RN11168} interprets as evidence of specialized production during Chimú times.  Tschauner’s excavations at the Pampa de Burros workshop in the Lambayeque Valley identified heavy use of vertical half-molds to produce jars, canteens or flasks, bottles, and ollas—virtually all of the shapes that were used by Chimú-era settlements in the surrounding valley \citep[182]{RN11168}.  He sees the local potters as relatively unskilled producers who relied on more adept makers of ceramic molds, and whose _market-oriented_ work could flexibly adapt to consumer preferences \citep[183-185]{RN11168}. \citet{RN11170} builds on this idea, noting that some of the variability seen in the Chimú assemblage could come from ceramic producers _mixing and matching_ molds for different vessel parts and adornments to produce new combinations.

For the Inca period, scholars acknowledge a degree of continuity in local potting practices, but also some significant changes.  At Tambo Real and La Viña in the Leche Valley, \citet{RN8875} notes the continuity of local practices for shaping vessels (press molds, paddle and anvil) and firing.  While local jar shapes persisted in Inca times, she observes the presence of a mold for an Inca flared jar neck \citep[345]{RN8875}, which could be added to jars that were shaped using local press molds.  Overall, Hayashida sees little evidence of Inca _retraining_ of Chimú potters.

Other Inca-Chimú workshops in the Jequetepeque Valley provide evidence that practices varied at different production locations. The surface collections of \citet[32-35]{RN11164} at Cañoncillo encountered dozens of mold fragments, most of which were for shaping the bottom and top hemispheres of ollas.  Donnan identified molds for shaping jars along a vertical axis, and for producing bottles, including the distinctive Inca _aryballoid_ shape.  Although he reports that no molds appeared for shaping handles or rims, there were molds for producing stirrup spouts, as well as for humans, birds, and other animals.  The production evidence from this site suggests that while some vessels were largely shaped using molds, some elements were hand-formed and showed variations suggesting the work of multiple potters.  At the Inca administrative center of Farfán, \citet[336-338]{RN11166} excavated a small workshop that produced large jars (_tinajas_) for brewing or storage, using the paddle-and-anvil technique.  Finally, the excavation of three workshop patios at El Algarrobal de Moro, a lower-order Chimú administrative center, encountered 25 mold fragments, including molds for a stirrup spout vessel and a miniature jar \citep[21-22]{RN11167}.

@@

Given the acknowledged challenges in classifying Chimú vessels, it is not surprising that recent ceramic studies have embraced technological analyses, including petrography \citep{RN11171}, neutron activation, X-ray diffraction, X-ray fluorescence, particle-induced X-ray emission \citep{RN11172}, and Mössbauer spectroscopy \citep{RN11173}.  Despite such work, \citet{RN11174} argue that the reconstruction of the chaîne opératoire for north coast blackwares remains incomplete.  It should be noted that many of the techniques used to reconstruct production practices rely on samples of potsherds that can be removed to the laboratory, and potentially altered or destroyed in the process of analysis.  More recently, researchers have brought non-destructive techniques such as CT scans \citep{RN11175} and linear morphometric analysis \citep{RN11159} to the study of museum collections.  This work emphasizes the variability found in different vessel categories, as well as the reconstruction of productive processes, including the use of molds to manufacture vessel bodies and other elements.

@@

These two questions can be addressed using three-dimensional data from well-preserved vessels, an approach that presents logistical challenges.  To date, \citet{RN11175} has used CT scans and X-ray radiography to study 16 stirrup-spout bottles from the Royal Museums of Art and History of Brussels and the Ethnography Museum of Geneva.  To produce the scans, the vessels had to be removed from the museums and transferred to hospitals where the necessary equipment could be accessed.  While generating high-resolution data, funding constraints and curatorial policies would make it difficult to conduct such research with large sample sizes.  As an alternative, portable high-resolution 3D scanning equipment can be brought into museums to scan collections in situ, and those data can be used for geometric morphometric analyses.

@@

For our broader study, we scanned #### prehispanic ceramic vessels from two collections that reflect the broader history of collecting and curating Chimú ceramics.  Our first collection was the Art and Art History Collection (AAHC), a modest sized collection that accumulated haphazardly at the University of Texas at Austin as numerous individuals donors gave the university artifacts that had little or no provenance information.  The ancient pottery now in the AAHC was previously curated by the Texas Memorial Museum, which transferred them to the department of Art and Art History in 2004 as the museum implemented new curatorial priorities. Those data presented here come from a sample of Chimú bottles (n=23) and jars (n=16) from the UT collection.  _**[Astrid:  How many Andean vessels overall in the collection, and how much Chimú?  Should we include a table of the samples here, with as much provenance info as we can find?]**_

The exploratory nature of this effort also raised ethical considerations regarding modern research with extant museum collections.  Archaeologists have intensified their work with museum collections in recent years, and the consideration of whether to conduct such research are as important as the logistical questions of how to accomplish the work.  The ethical principles of the Society for American Archaeology highlight the values of stewardship and public responsibility—it is important for archaeologists to care for the material record, and also to share information about it with the public—identifying “public education and outreach” as significant.  From this perspective, the non-destructive and museum-based work of 3D scanning presented little risk to the artifacts being analyzed, and our research was designed to be available to the public.  When working with the AAHC vessels, we developed an educational component, including a small class of UT undergraduates in the lab-based collection of linear measurements of the vessels being scanned.  That supervised research provided a diverse group of undergraduates with artifact analysis, while also teaching them about the loss of heritage and scientific knowledge that accompanies the collection of antiquities.

While the careful study and public reporting of museum collections is consistent with professional ethics in Americanist archaeology, the SAA also identifies a potential ethical concern in doing so:  the commercialization of the archaeological record.  The SAA clearly states that “Archaeologists should therefore carefully weigh the benefits to scholarship of a project against the costs of potentially enhancing the commercial value of archaeological objects. Whenever possible they should discourage, and should themselves avoid, activities that enhance the commercial value of archaeological objects, especially objects that are not curated in public institutions, or readily available for scientific study, public interpretation, and display.”   As described above, Chimú blackware pottery is an artifact class that was heavily collected in the past, and it is openly sold today by galleries and internet vendors.  As \citet[22-23]{RN11153} notes, huaqueros, dealers, and collectors are no longer interested in trafficking Andean ceramics, especially mass-produced wares like those of the Chimú Empire.  Our work with publicly-held collections poses little risk of increasing the present commercial value of Chimú ceramics, while its public discussion of the losses that accompany undocumented excavations and illicit collection constitutes a benefit that outweighs any associated risks.

@@

Linear measurements from the sample of jars and bottles included vessel height, rim thickness, and the diameter of each vessel at its orifice, neck, belly, and base (Figure \ref{fig:linear}).  Of these metrics, vessel height, base diameter, and neck diameter did not differ significantly between the two samples ([Supplementary Materials, Ch. 1](https://seldenlab.github.io/peru.bottle.jar/linear-chim%C3%BA-vessel-metrics.html)).  Overall, the jars and bottles in the sample overlap in some key dimensions, with the bottles showing greater overall size variations, which is consistent with production that uses a single set of molds.  The other three sets of linear measurements demonstrated significant differences between the jars and bottles for orifice diameter, belly diameter, and rim thickness ([Supplementary Materials, Ch. 1](https://seldenlab.github.io/peru.bottle.jar/linear-chim%C3%BA-vessel-metrics.html)).  Given the different production scenarios emerging from archaeological field analyses---a combination of hand-shaping and use of whole vessel or elemental molds, the significant differences in orifice and belly diameters may indicate that the kinds of bottles and jars in our sample were produced as distinct shapes, rather than by _mixing and matching_ different molds to cater to changing consumer tastes.  The differences in rim thickness suggests that distinct production practices were used when producing bottles and jars.

@@

Following transfer to Dx, each mesh will be subjected to an additional quality check to eliminate non-manifold poly-vertices, folded poly-faces, dangling poly-faces, small clusters, small poly-faces, non-manifold poly-faces, crossing poly-faces, and small tunnels. Due to the paucity of homologous landmarks on cultural artefacts \citep{RN8521}, reference geometry will be constructed around each vessel in a manner that yields a replicable configuration of nine landmarks, and 46 equidistant semilandmarks along the widest vessel profile, with notable similarities to previous landmark configurations \citep{RN7039,RN7925,RN8071,RN8361,RN8967,RN7588}, which largely follow \citet{RN5700}.

@@

Landmark data were aligned to a global coordinate system \citep{RN8477,RN7502,RN11622,RN11623,RN11563}, achieved through generalised Procrustes superimposition \citep{RN11138,RN478,RN1646} in R using the `geomorph` and `RRPP` packages \citep{RN1655,RN11775,RN11530,RN1774,RN9565}. Procrustes superimposition translates and rotates the coordinate data to allow for comparisons among objects, while also scaling each biface using unit-centroid size---the square root of the sum of squared distances from each landmark to the specimen’s centroid \citep{RN11139,RN11140,RN5622,RN478}. The `geomorph` package uses a partial Procrustes superimposition that projects the aligned specimens into tangent space subsequent to alignment in preparation for the use of multivariate methods that assume linear space (Figure \ref{fig:gpa}) \citep{RN11141,RN11142,RN1646,RN11563}.

@@
